draft

### 2025-05-25 draft.tex
Updates from Overleaf

\subsection{The current state of bioimage analysis}
\label{sec:org8f5b33d}
The most widely used software for processing high-throughput biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}. It can be used by experimental biologists with scant programming experience, for whom it has been a boon. By contrast, it may not suit the needs of computer-savvy scientists building high throughput pipelines comprised of multiple tools. CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance. For example, complementing CellProfiler with other image analysis tools is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces. Lastly, CellProfiler depends on many Python and Java packages, increasing the likelihood of dependency conflicts. Containers mitigate this problem, but not without increased complexity and caveats.

@@

%First, we demonstrated the validity and capabilities of cp\_measure in cases that would normally require image analysis knowledge and a considerable amount of time. First, we validate cp\_measure features versus CellProfiler results with a subset of the JUMP dataset \citep{chandrasekaranJUMPCellPainting2023}. Then we showcase cases in which cp\_measure is a more practical choice to process microscopy data: First 3D images of astrocytes and then spatial transcriptomics. These use-cases demonstrate its widespread suitability for different types of problems. 
\subsection{Reproducing CellProfiler measurements}
\label{sec:org09b0cd2}

@@

We first tested whether cp\_measure features match the original CellProfiler features, for a representative set of 300 images. These correspond to 150 perturbations from the JUMP Cell Painting dataset \citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for each feature. We segmented images to delineate the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to cp\_measure for feature extraction. Next, we applied cp\_measure on these masks and the original images. Finally, we mapped the features from cp\_measure to CellProfiler and calculated a linear fit for each matched feature, plus the \(R^2\) value, indicating how well the correlation fits a linear slope.

We find that most features had an $R^2$ correlation larger than 90\% (Figure \ref{fig:cp_vs_cpmeasure}). We examined several individual features, typically finding straight lines demonstrating the recapitulation of features from our implementation. For a small set of features, some data points fall off the diagonal, which hint that some edge cases are treated differently by either tool; we plan to investigate these further as part of a larger project to implement unit testing in CellProfiler. These results indicate that cp\_measure is not likely to produce massively different profiles than those calculated by CellProfiler.
\subsection{Results}
\label{sec:orge5b5c6b}
\subsubsection{Astrocyte 3D nuclei data}
\label{sec:org447090b}

@@

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/example_shap.svg}
\caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astrocyte nuclei image and masks. The 3D images were projected over the z-axis, taking the maximum value across the z-stack. \textbf{Bottom panel.} SHAP values of the most important features for classifying the day of differentiation (out of three days). Each point represents a single 3D image containing multiple nuclei, the magnitude of SHAP values indicate how important those features are for the classifier. Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value. Shape features refer to the nucleus and intensity features refer to the DNA staining. The test data accuracy is shown in bold.}
\end{figure}

Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the SHAP values of a classifier trained on our features. The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover.
\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
A key advantage of providing these widely used features with proven usefulness as a stand-alone Python package is how easy this makes integration into diverse analytical workflows. To demonstrate this, we have developed a module based on cp\_measure for the popular spatial omics analysis toolbox Squidpy \citep{pallaSquidpyScalableFramework2022}.
Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E stained or fluorescently labeled.

While not the typical application domain of CellProfiler, the recent proliferation of black-box foundation models trained solely on morphological data \citep{Chen2024-ix, Xu2024-mn} highlights morphology as a highly informative and predictive modality. However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment. In contrast, using cp\_measure to featurize such images will yield more interpretable features.

However, directly applying CellProfiler to spatial datasets is not convenient since it imposes a nested structure on input data and also cannot handle extremely large images since the full image is loaded into RAM for processing. Since spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screening data, easily 500.000 cells per image, we used cp\_measure to develop a workflow tailored to such data. For this, Squidpy leverages the powerful data format SpatialData \citep{Marconato2024-SpatialData} in which images are stored as zarr objects. This allows us to parallelize the feature extraction with cp\_measure at the mask level, for example individual cells, streaming batches of cells across computational cores. This approach enables efficient computation even on large-scale datasets, a feat not achievable with standard CellProfiler. Furthermore, this establishes morphological profiling within the scverse ecosystem \citep{Virshup2023-scverse}, exposing it to the wider single-cell community.

To illustrate the value of this workflow, we evaluated its usefulness in identifying morphologically similar regions within two human breast cancer sections profiled with the spatial transcriptomics technology Xenium by 10x Genomics \citep{10x-Genomics2023-el}. We generated a tessellating pattern of hexagonal masks across both images, of which XXXXXX were found to overlay the tissue. These masks were then featurised with the described cp\_measure-based workflow. We first preprocessed the features as described in 
\citep{serranoReproducibleImagebasedProfiling2025}, then integrated both samples using Harmony \citep{Korsunsky2019-Harmony}, and finally clustered them with the Leiden algorithm through Scanpy \citep{wolfSCANPYLargescaleSinglecell2018}.

For each cluster, we extracted the cells within its hexagons and calculated the proportions of cell types predicted by the spatial gene-expression data. A visual comparison of panels A and B in Figure \ref{fig:spatial_omics} shows that morphology-only clustering largely recapitulates the underlying tissue architecture. Panel C further reveals that some clusters have distinct cell-type compositions: for example, Cluster XXX is dominated by fibroblasts, whereas adipocytes appear almost exclusively in Cluster XXX.

\begin{figure}[htbp]
\centering
\includegraphics[width=.99\linewidth]{./figs/fig_4_placeholder.png}
\caption{\label{fig:spatial_omics}{[}PLACEHOLDER] Spatial omics analysis.}
\end{figure}

These results demonstrate that our cp\_measure–powered workflow not only scales to the high cell counts typical of spatial datasets but also produces biologically meaningful, interpretable morphological clusters that align with - and complement - transcriptomic phenotypes. By embedding this capability within the scverse ecosystem via Squidpy, we enable researchers to seamlessly integrate morphology into their spatial omics analyses, unlocking new insights into tissue organization and cell–cell interactions.

@@

Already, Cellprofiler features have been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}. By association, we foresee cp\_measure being useful for a broad scientific community beyond image-based profiling.
\section{Future work}
\label{sec:org5cdbb12}
We propose in future work to make it possible for CellProfiler to import cp\_measure, which offers several benefits. It would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements. It would also mean that any changes made in cp\_measure propagate to CellProfiler, benefiting the community for which it was originally developed.

@@

\bibliographystyle{icml2025}
\bibliography{bibliography}
\section{Appendix}
\label{sec:orgdd18dd8}
\subsection{Methods}
\label{sec:orgb3e9382}
\subsubsection{Data and software}
\label{sec:orgbda0ae2}
The code for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}. All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://github.com/afermg/2025\_cpmeasure/}. The datasets we produced for this work are available on Zenodo, and the latest version can be found on \url{https://zenodo.org/records/15390631/latest}.
\subsubsection{Data and software}
\label{sec:acknowledgements}

>>>>

\subsection{The current state of bioimage analysis}
\label{sec:org8f5b33d}
The most widely used software for processing high-throughput biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}. It can be used by experimental biologists with scant programming experience, for whom it has been a boon. By contrast, it may not suit the needs of computer-satvvy scientists building high throughput pipelines comprised of multiple tools. CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance. For example, complementing CellProfiler with other image analysis tools is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces. Lastly, CellProfiler depends on many Python and Java packages, increasing the likelihood of dependency conflicts. Containers mitigate this problem, but not without increased complexity and caveats.

@@

%First, we demonstrated the validity and capabilities of cp\_measure in cases that would normally require image analysis knowledge and a considerable amount of time. First, we validate cp\_measure features versus CellProfiler results with a subset of the JUMP dataset \citep{chandrasekaranJUMPCellPainting2023}. Then we showcase cases in which cp\_measure is a more practical choice to process microscopy data: First 3D images of astrocytes and then spatial transcriptomics. These use-cases demonstrate its widespread suitability for different types of problems. 
\subsection{Reproducing CellProfiler measurements}
\label{sec:org09b0cd2}
We first tested whether cp\_measure features match the original CellProfiler features, for a representative set of 300 images. These correspond to 150 perturbations from the JUMP Cell Painting dataset \citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for each feature. We segmented images to delineate the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to cp\_measure for feature extraction. Next, we applied cp\_measure on these masks and the original images. Finally, we mapped the features from cp\_measure to CellProfiler and calculated a linear fit for each matched feature, plus the \(R^2\) value, indicating how well the correlation fits a linear slope.

We find that most features had an $R^2$ correlation larger than 90\% (Figure \ref{fig:cp_vs_cpmeasure}). We examined several individual features, typically finding straight lines demonstrating the recapitulation of features from our implementation. For a small set of features, some data points fall off the diagonal, which hint that some edge cases are treated differently by either tool; we plan to investigate these further as part of a larger project to implement unit testing in CellProfiler. These results indicate that cp\_measure is not likely to produce massively different profiles than those calculated by CellProfiler.

@@

\subsection{Results}
\label{sec:orge5b5c6b}
\subsubsection{Astrocyte 3D nuclei data}
\label{sec:org447090b}

@@

Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the SHAP values of a classifier trained on our features. The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover.

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/example_shap.svg}
\caption{\label{fig:astrocytes}\textbf{A,B.} Example pair of astrocyte nuclei image and masks. The 3D images were projected over the z-axis, taking the maximum value across the z-stack. \textbf{Bottom panel.} SHAP values of the most important features for classifying the day of differentiation (out of three days). Each point represents a single 3D image containing multiple nuclei, the magnitude of SHAP values indicate how important those features are for the classifier. Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value. Shape features refer to the nucleus and intensity features refer to the DNA staining. The test data accuracy is shown in bold.}
\end{figure}

\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
To demonstrate the broad applicability of cp\_measure, we developed a morphology featurizer with it at its core for the popular spatial omics analysis toolbox Squidpy \citep{pallaSquidpyScalableFramework2022}.
Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E (hematoxylin and eosin) stained or fluorescently labeled.

The recent proliferation of black-box foundation models trained solely on morphological data \citep{Chen2024-ix, Xu2024-mn} highlights morphology as a highly informative and predictive modality. However, the features these models are typically not interpretable, preventing direct biological assessment. In contrast, using cp\_measure to featurize such images will yield more interpretable features.

However, directly applying CellProfiler to spatial datasets is not convenient since it imposes a nested structure on input data and also cannot handle extremely large images since the full image is loaded into RAM for processing. Since spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screening data, easily 500.000 cells per image, the standalone nature of cp\_measure allowed us to tailor a workflow to such data. For this, Squidpy leverages the powerful data format SpatialData \citep{Marconato2024-SpatialData} in which images are stored as zarr objects. This allows us to parallelize the feature extraction with cp\_measure at the mask level, for example individual cells, streaming batches of cells across computational cores. This approach enables efficient computation even on large-scale datasets, a feat not achievable with standard CellProfiler. Furthermore, this establishes morphological profiling within the scverse ecosystem \citep{Virshup2023-scverse}, exposing it to the wider single-cell community.

To illustrate the value of this workflow, we evaluated its usefulness in identifying morphologically similar regions within two human breast cancer sections profiled with the spatial transcriptomics technology Xenium by 10x Genomics \citep{10x-Genomics2023-el}. We generated a tessellating pattern of hexagonal masks across both images, of which XXXXXX were found to overlay the tissue. These masks were then featurized with the described cp\_measure-based workflow. We first preprocessed the features as described in 
\citep{serranoReproducibleImagebasedProfiling2025}, then integrated both samples using Harmony \citep{Korsunsky2019-Harmony}, and finally clustered them with the Leiden algorithm through Scanpy \citep{wolfSCANPYLargescaleSinglecell2018}.

For each cluster, we extracted the cells within its hexagons and calculated the proportions of cell types predicted by the spatial gene-expression data. A visual comparison of panels A and B in Figure \ref{fig:spatial_omics} shows that morphology-only clustering largely recapitulates the underlying tissue architecture. Panel C further reveals that some clusters have distinct cell-type compositions: for example, cluster 2 is dominated by fibroblasts, whereas adipocytes appear almost exclusively in cluster 5-7.

\begin{figure}[htbp]
\centering
\includegraphics[width=.99\linewidth]{./figs/fig_4_spatial.png}
\caption{\label{fig:spatial_omics}{}Morphology-driven tumor subregion discovery. \textbf{Panel A}: Spatial map of cell types inferred from transcriptomics. \textbf{Panel B}: Morphological subregions obtained by applying Leiden clustering to image-derived feature embeddings.. \textbf{Panel C}: Bar chart showing, for each morphological cluster, the fraction of cells assigned to each cell type (color legend as shown). Of note is that the cell segmentation in the top-left quarter of the image seems to have performed poorly, since no cells were annotated, whereas the morphological clusters cover these regions.}
\end{figure}

These results demonstrate how cp\_measure can now easily be used outside of its traditional domain. Using it to featurize extremely large images, we show that it produces biologically meaningful morphological clusters that align with - and complement - transcriptomic phenotypes. By embedding this capability within the scverse ecosystem via Squidpy, we enable researchers to seamlessly integrate morphology into their spatial omics analyses, unlocking new insights into tissue organization and cell–cell interactions.

@@

Already, Cellprofiler features have been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}. By association, we foresee cp\_measure being useful for a broad scientific community beyond image-based profiling.
\section{Future work}
\label{sec:org5cdbb12}
We propose to make cp\_measure an imported dependency for CellProfiler, offering several benefits. It would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements. It would also mean that any changes made in cp\_measure propagate to CellProfiler, benefiting the community for which it was originally developed.

@@

\bibliographystyle{icml2025}
\bibliography{bibliography}
\section{Appendix}
\label{sec:orgdd18dd8}
\subsection{Methods}
\label{sec:orgb3e9382}
\subsubsection{Data and software}
\label{sec:orgbda0ae2}
The code for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}. All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://github.com/afermg/2025\_cpmeasure/}. The datasets we produced for this work are available on Zenodo, and the latest version can be found on \url{https://zenodo.org/records/15390631/latest} and \url{https://zenodo.org/records/15505477/latest}.
\subsubsection{Acknowledgements}
\label{sec:acknowledgements}
